---
title: "Class 17"
format: pdf
execute:
  working-directory: /Users/yasoohaitham
---

```{r}
 getwd()
```




```{r}
path.expand("~/Downloads")
```



```{r}
list.files(path.expand("/Users/yasoohaitham/Downloads"))
```



```{r}
samples <- data.frame(
  row.names = c("SRR2156848", "SRR2156849", "SRR2156850", "SRR2156851"),
  condition = c("control","control","treated","treated")
)
```


```{r}
library(tximport)

quant_dirs <- c(
  "/Users/yasoohaitham/Downloads/SRR2156848_quant/abundance.h5",
  "/Users/yasoohaitham/Downloads/SRR2156849_quant/abundance.h5",
  "/Users/yasoohaitham/Downloads/SRR2156850_quant/abundance.h5",
  "/Users/yasoohaitham/Downloads/SRR2156851_quant/abundance.h5"
)

txi <- tximport(
  files = quant_dirs,
  type  = "kallisto",
  txOut = TRUE
)

names(txi)
```


```{r}
library(DESeq2)

dds <- DESeqDataSetFromTximport(
  txi      = txi,
  colData  = samples,
  design   = ~ condition
)

dds
```

```{r}
vsd <- varianceStabilizingTransformation(dds)
vst_mat <- assay(vsd)
```
```{r}
nonzero <- apply(vst_mat, 1, var) > 0
vst_filt <- vst_mat[nonzero, ]
```

```{r}
pca <- prcomp(t(vst_filt), scale = TRUE)
pca
```
```{r}
df <- as.data.frame(pca$x)
df$sample <- rownames(df)
df$condition <- samples$condition
df
```




```{r}
list.files(path.expand("~/Downloads"), pattern = "SRR")
```



```{r}
samples <- data.frame(
  sample = c("SRR2156848", "SRR2156849", "SRR2156850", "SRR2156851"),
  condition = c("control", "control", "treated", "treated")
)
samples
```



```{r}
quant_dirs <- c(
  "~/Downloads/SRR2156848_quant/quant.sf",
  "~/Downloads/SRR2156849_quant/quant.sf",
  "~/Downloads/SRR2156850_quant/quant.sf",
  "~/Downloads/SRR2156851_quant/quant.sf"
)

quant_dirs
```

```{r}
list.files("/Users/yasoohaitham/Downloads", pattern="SRR", full.names=TRUE)
```

```{r}
quant_dirs <- c(
  "/Users/yasoohaitham/Downloads/SRR2156848_quant/quant.sf",
  "/Users/yasoohaitham/Downloads/SRR2156849_quant/quant.sf",
  "/Users/yasoohaitham/Downloads/SRR2156850_quant/quant.sf",
  "/Users/yasoohaitham/Downloads/SRR2156851_quant/quant.sf"
)

file.exists(quant_dirs)
```



```{r}
library(DESeq2)

samples <- data.frame(
  row.names = c("SRR2156848", "SRR2156849", "SRR2156850", "SRR2156851"),
  condition = c("control", "control", "treated", "treated")
)

dds <- DESeqDataSetFromTximport(
  txi      = txi,
  colData  = samples,
  design   = ~ condition
)

dds
```



```{r}
df <- as.data.frame(pca$x)
df$sample <- rownames(df)
df$condition <- c("control","control","treated","treated")

head(df)
```


```{r}
setwd("~/Downloads")
```

setwd("~/Downloads")

## Core Unix commands
Most unix commands have super short names, which makes them quick to type but annoying to learn. Major file system related commands include:

pwd: Print working directory 
ls: list files and directories 
cd: Change Directory
mkdir: Make a new directory rm: remove files and directory (delete)
ср: Copy files or dirs (source > destination) 
mv: Move files or dirs (basically re-name)
nano: A wee text editor (very basic but always available)

curl: Download files from web or ftp 
wet: Download files from the web 
tar -zxvf: UnTar (unpackage) Tar archive files 
gunzip: UnZip files
$PATH: The places (dirs) to look for programs


## AWS EC2 Instance


ssh -i ~/Downloads/bimm143_ya.pem ubuntu@ec2-34-222-197-175.us-west-2.compute.amazonaws.com
ssh -i "bioinf_yasmeen.pem" ubuntu@ec2-34-222-197-175.us-west-2.compute.amazonaws.com
ssh -i  ~/Downloads/bioinf_yasmeen.pem ubuntu@ec2-34-222-197-175.us-west-2.compute.amazonaws.com

Secure Copy files between machines, in this case from our instance to our laptop
scp -i ~/Downloads/bioinf_yasmeen.pem ubuntu@ec2-16-145-48-135.us-west-2.compute.amazonaws.com:~/results.tsv . 



## Class 17 Instance 
ssh -i  ~/Downloads/bioinf_yasmeen.pem ubuntu@ec2-34-222-197-175.us-west-2.compute.amazonaws.com

export KEY=~/Downloads/bioinf_yasmeen.pem

export SERVER=ubuntu@ec2-34-210-54-74.us-west-2.compute.amazonaws.com

ssh -i $KEY $SERVER
scp -r -i $KEY $SERVER:~/*_quant .cd ~/Downloads

>Q1: What command prints the first lines of a FASTQ file?

head SRR2156848_1.fastq

>Q2: How to count the number of reads?

grep -c "^@" SRR2156848_1.fastqhead SRR2156848_1.fastq




```{r}
library(ggplot2)

df <- as.data.frame(pca$x)
df$sample <- rownames(df)
df$condition <- c("control","control","treated","treated")

```



```{r}
ggplot(df, aes(PC1, PC2, color=condition)) +
  geom_point(size=4) +
  theme_minimal() +
  labs(title="PC1 vs PC2")

```


```{r}
ggplot(df, aes(PC1, PC3, color=condition)) +
  geom_point(size=4) +
  theme_minimal() +
  labs(title="PC1 vs PC3")
```

```{r}
ggplot(df, aes(PC2, PC3, color=condition)) +
  geom_point(size=4) +
  theme_minimal() +
  labs(title="PC2 vs PC3")

```


```{r}
plot(pca$x[,1], pca$x[,3],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC3", pch=16)
```



```{r}
plot(pca$x[,2], pca$x[,3],
     col=c("blue","blue","red","red"),
     xlab="PC2", ylab="PC3", pch=16)

```

```{r}
plot(pca$x[,2], pca$x[,3],
     col=c("blue","blue","red","red"),
     xlab="PC2", ylab="PC3", pch=16)
```

```{r}
library(ggplot2)
library(ggrepel)

mycols <- c("blue","blue","red","red")

ggplot(pca$x) +
  aes(PC1, PC2, label=rownames(pca$x)) +
  geom_point(col = mycols) +
  geom_text_repel(col = mycols) +
  theme_bw()
```


```{r}
ggplot(pca$x) +
  aes(PC1, PC3, label=rownames(pca$x)) +
  geom_point(col = mycols) +
  geom_text_repel(col = mycols) +
  theme_bw()
```


```{r}
ggplot(pca$x) +
  aes(PC2, PC3, label=rownames(pca$x)) +
  geom_point(col = mycols) +
  geom_text_repel(col = mycols) +
  theme_bw()
```



















### NOTE about graphs: lab screenshot used the full Ensembl transcriptome where i had to use the MINI transcriptome. However, the good news is that it does NOT the PCA pattern! So the pattern shoudl still be corrrect. 




