---
title: "Class 08: Breast Cancer Mini Project"
author: "Yasmeen Al Shakarchi (PID: A18116120)"
format: pdf
toc: true
---

## Background 

In today's class we will be employing all the R techniques for data analysis that we have learned thus far for - inlcuding the machine learning methods of clustering and PCA - to analyze real breast cancer biopsy

## Data Import

The data is in CSV format: 

```{r}
wisc.df <- read.csv("WisconsinCancer.csv", row.names = 1)
```

wee peak at the data
```{r}
head(wisc.df, 3)
```
> Q1. How many obervations are in this dataset?

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```



```{r}
nrow(wisc.df)
```


> Q2. How many of the observations have a malignant diagnosis? 

```{r}
sum(wisc.df$diagnosis == "M")
```
```{r}
table(wisc.df$diagnosis)

```

> Q3. How many variables/features in the data are suffixed with _mean? 

```{r}
length(grep("_mean",colnames(wisc.df)))
```

```{r}
colnames(wisc.df)
```

We need to remove the `dignosis` column before we do any further analysis of this dataset- we don't want to pass this to PCA etc. We will save it as a separate wee vector that we can use later to compare our findings to those of experts. 

```{r}
wisc.data <- wisc.df[,-1]
diagnosis <- wisc.df$diagnosis
```

## Principal Component Analysis (PCA)

The main function in base R is called `prcomp()` we will use the optional argument `scale = TRUE` here as the data columns/features/dimensions are on very different scales in the original data set.

```{r}
wisc.pr <- prcomp(wisc.data, scale= T)
```

```{r}
head(wisc.pr$x)
```

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC2, col=diagnosis) +
  geom_point()
```

```{r}
summary(wisc.pr)
```

> Q4. From your results, what proportion of the original variance is captured by the first principal component (PC1)?

Answer: 0.4427

> Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

Answer: PC3

> Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

Answer: PC7

> Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

```{r}
biplot(wisc.pr)
```

Answer: This plot is very messy and not easy to read at all. The lines overlap and it is not comprehensible. 



### Scree plot

```{r}
pr.var <- wisc.pr$sdev^2
```


```{r}
pve <- pr.var / sum(pr.var)

plot(c(1, pve), 
     xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), 
     type = "o")
```

```{r}
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```


```{r}
library(factoextra)
fviz_eig(wisc.pr, addlabels = TRUE)
```
> Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

```{r}
library(ggplot2)

ggplot(wisc.pr$x) +
  aes(PC1, PC3, col=diagnosis) +
  geom_point()
```
Answer: The plot is tighter and PC3 explains less variance than PC2 (Less variation). 

> Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean? This tells us how much this original feature contributes to the first PC. Are there any features with larger contributions than this one?


```{r}
wisc.pr$rotation["concave.points_mean", "PC1"]
```


```{r}
abs(wisc.pr$rotation[,1])
```


Answer: concave.points_mean contributes fairly strongly to PC1 ( neg (-)), the answer here was –0.2608. Yes lots of others have larger absolute loadings. 



## 4. Hierarchial clustering 

The goal of this section is to do heierarchical clustering of the original data to see if there is any obvious grouping into malignant and benign clusters. 

In short, these results are not good!

First we will scale our `wisc.data` then calculate a distance matrix, then pass to `hclust()`: 

```{r}
wisc.dist <- dist( scale(wisc.data))
wisc.hclust <- hclust(wisc.dist)
plot(wisc.hclust)
```

```{r}
plot(wisc.hclust)
abline(h = 19, col="red", lty=2)
```
> Q10. Using the plot() and abline() functions, what is the height at which the clustering model has 4 clusters?

```{r}
plot(wisc.hclust, labels = FALSE, hang = -1, cex = 0.3)
abline(h = 7, col = "red", lty = 2)
```

Answer: At h = 19. 


  > Q12. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.


Answer: My favorite results come from "average" because 


```{r}
wisc.hclust.clusters <- cutree(wisc.hclust, k=2)
table(wisc.hclust.clusters)
```



## Combining methods 

The idea here is that I can take my new variables (i.e. the scores on the pCs `wisc.pr$x`) that are better descriptions of the data-set than the original features (i.e. the 30 columns in `wisc.data`) and use these as a basis for clustering.

```{r}
pc.dist <- dist(wisc.pr$x[, 1:3])
wisc.pr.hclust <- hclust(pc.dist, method= "ward.D2")
plot(wisc.pr.hclust)
```


```{r}
grps <- cutree(wisc.pr.hclust, k=2)
table(grps)
```

```{r}
table(diagnosis)
```
 

```{r}
wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=2)
```

> Q13. How well does the newly created hclust model with two clusters separate out the two “M” and “B” diagnoses?

```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```

Answer: It clusters it better because there is two clusters separates the diagnoses ( one fouces on B and other on M). 

I can now run `table()` with both my clustering `grps` and the expertn `diagnosis`

```{r}
table(grps, diagnosis)
```
Our clustering "1" has 179 "M" diagnosis 
or cluster "2" has 333 "B" diagnosis


179 TP 
24 FP
333 TN
33 FN

> Q14. How well do the hierarchical clustering models you created in the previous sections (i.e. without first doing PCA) do in terms of separating the diagnoses? Again, use the table() function to compare the output of each model (wisc.hclust.clusters and wisc.pr.hclust.clusters) with the vector containing the actual diagnoses.

```{r}
table(wisc.hclust.clusters, diagnosis)
```


```{r}
table(wisc.pr.hclust.clusters, diagnosis)
```

Answer: The hierarchical clustering model without pr did a poor job separating the diagnoses ebc ause it mixed both diagnoses toghther. Where the hierarchical clustering model with pr did a poor job separating the diagnoses to create two clusters that more clearly separated benign and malignant samples ( better diagnostic separation).




> Q15. OPTIONAL: Which of your analysis procedures resulted in a clustering model with the best specificity? How about sensitivity?

Sensitivity: TP/(TP+FN)
```{r}
179/(179+33)
```


Specificity:  TN/(TN+FP)
```{r}
333/(333+24)
```
> 

## Prediction

We can use our PCA model for predicition of new un-seen cases: 

```{r}
url <- "https://tinyurl.com/new-samples-CSV"
new <- read.csv(url)
npc <- predict(wisc.pr, newdata=new)
npc
```


```{r}
plot(wisc.pr$x[,1:2], col=grps)
points(npc[,1], npc[,2], col="blue", pch=16, cex=3)
text(npc[,1], npc[,2], c(1,2), col="white")
```


> Q16. Which of these new patients should we prioritize for follow up based on your results


Answer: Based on these results the patient that we should prioritize is patient 2. This is because it is near the black cluster (malignant samples).
