---
title: "Class 12 lab session"
author: "Yasmeen Al Shakarchi"
date: "2026-02-16"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Section 1. Proportion of G/G in population

Downloaded a csv file from Ensemble < https://useast.ensembl.org/Homo_sapiens/Variation/Sample?db=core;r=17:39891511-39891622;v=rs8067378;vdb=variation;vf=959672880#373531_tablePanel >


Here we read this CVS file


```{r}
mxl <- read.csv(("373531-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv"))
head(mxl)
```

```{r}
table(mxl$Genotype..forward.strand.)
```



```{r}
table(mxl$Genotype..forward.strand.) / nrow(mxl) *100
```


Now let's look at a different population. I picked the GBR. 

```{r}
gbr <- read.csv("373522-SampleGenotypes-Homo_sapiens_Variation_Sample_rs8067378.csv")
head(gbr)
```
Find the proportion of G/G



```{r}
round(table(gbr$Genotype..forward.strand.) / nrow(gbr) *100, 2)
```

This variant that is associated with childhood asthma is more frequent in the GBR population than the MKL population.


Let's now dig into this further.


Section 4: Population Scale Analysis [HOMEWORK]
One sample is obviously not enough to know what is happening in a population. You are interested in assessing genetic differences on a population scale. 

How many samples do we have?

```{r}
expr <- read.table("rs8067378_ENSG00000172057.6.txt")
head(expr)
```
 
```{r}
nrow(expr)
```


> Q13. Sample size?

```{r}
table(expr$geno)
```


> Q12. Median? 

```{r}
tapply(expr$exp, expr$geno, median)
```



```{r}
library(ggplot2)
```


Lets make a boxplot


```{r}
ggplot(expr) + aes(geno, exp, fill=geno) +
  geom_boxplot(notch = TRUE)
```




>Q14: Generate a boxplot with a box per genotype, what could you infer from the relative expression value between A/A and G/G displayed in this plot? Does the SNP effect the expression of ORMDL3?

Answer: Yes SNP does affect ORMDL3 expression. Having a G/G in this location will cause reduced expression (A/A contains the highest expression/ higher expression). 


