---
title: "Class 18"
author: "Yasmeen Al Shakarchi (PID: A18116120)"
format: pdf
---


## Background
Pertussis (a.k.a. whooping cough) is a common lung infection caused by bacteria B. Pertussis.

This can infect all ages but is most sever for those under 1 year of age.

The CDC track the number of reported cases in US:We can "scrape" this data with the **datapasta** package


```{r}
data <- data.frame(
      stringsAsFactors = FALSE,
           check.names = FALSE,
          `cdc.<-.data.frame(` = c("year = c(","1922L,","1923L,","1924L,","1925L,",
                                   "1926L,","1927L,","1928L,","1929L,",
                                   "1930L,","1931L,","1932L,","1933L,","1934L,",
                                   "1935L,","1936L,","1937L,","1938L,",
                                   "1939L,","1940L,","1941L,","1942L,","1943L,",
                                   "1944L,","1945L,","1946L,","1947L,",
                                   "1948L,","1949L,","1950L,","1951L,","1952L,",
                                   "1953L,","1954L,","1955L,","1956L,",
                                   "1957L,","1958L,","1959L,","1960L,","1961L,",
                                   "1962L,","1963L,","1964L,","1965L,",
                                   "1966L,","1967L,","1968L,","1969L,","1970L,",
                                   "1971L,","1972L,","1973L,","1974L,",
                                   "1975L,","1976L,","1977L,","1978L,","1979L,",
                                   "1980L,","1981L,","1982L,","1983L,",
                                   "1984L,","1985L,","1986L,","1987L,","1988L,",
                                   "1989L,","1990L,","1991L,","1992L,",
                                   "1993L,","1994L,","1995L,","1996L,","1997L,",
                                   "1998L,","1999L,","2000L,","2001L,",
                                   "2002L,","2003L,","2004L,","2005L,","2006L,",
                                   "2007L,","2008L,","2009L,","2010L,",
                                   "2011L,","2012L,","2013L,","2014L,","2015L,",
                                   "2016L,","2017L,","2018L,","2019L,",
                                   "2020L,","2021L,","2022L,","2023L,","2024L,",
                                   "2025L","),","cases = c(","107473,",
                                   "164191,","165418,","152003,","202210,",
                                   "181411,","161799,","197371,","166914,",
                                   "172559,","215343,","179135,","265269,",
                                   "180518,","147237,","214652,","227319,",
                                   "103188,","183866,","222202,","191383,","191890,",
                                   "109873,","133792,","109860,","156517,",
                                   "74715,","69479,","120718,","68687,",
                                   "45030,","37129,","60886,","62786,","31732,",
                                   "28295,","32148,","40005,","14809,",
                                   "11468,","17749,","17135,","13005,","6799,",
                                   "7717,","9718,","4810,","3285,","4249,",
                                   "3036,","3287,","1759,","2402,","1738,",
                                   "1010,","2177,","2063,","1623,","1730,",
                                   "1248,","1895,","2463,","2276,","3589,",
                                   "4195,","2823,","3450,","4157,","4570,",
                                   "2719,","4083,","6586,","4617,","5137,",
                                   "7796,","6564,","7405,","7298,","7867,",
                                   "7580,","9771,","11647,","25827,",
                                   "25616,","15632,","10454,","13278,","16858,",
                                   "27550,","18719,","48277,","28639,",
                                   "32971,","20762,","17972,","18975,","15609,",
                                   "18617,","6124,","2116,","3044,","7063,",
                                   "22538,","21996",")",")")
        )
```



```{r}
cdc <- tibble::tribble(
  ~year, ~cases,
  1922, 107473,
  1923, 164191,
  1924, 165418,
  1925, 152003,
  1926, 202210,
  1927, 181411,
  1928, 161799,
  1929, 197371,
  1930, 166914,
  1931, 172559,
  1932, 215343,
  1933, 179135,
  1934, 265269,
  1935, 180518,
  1936, 147237,
  1937, 214652,
  1938, 227319,
  1939, 103188,
  1940, 183866,
  1941, 222202,
  1942, 191383,
  1943, 191890,
  1944, 109873,
  1945, 133792,
  1946, 109860,
  1947, 156517,
  1948, 74715,
  1949, 69479,
  1950, 120718,
  1951, 68687,
  1952, 45030,
  1953, 37129,
  1954, 60886,
  1955, 62786,
  1956, 31732,
  1957, 28295,
  1958, 32148,
  1959, 40005,
  1960, 14809,
  1961, 11468,
  1962, 17749,
  1963, 17135,
  1964, 13005,
  1965, 6799,
  1966, 7717,
  1967, 9718,
  1968, 4810,
  1969, 3285,
  1970, 4249,
  1971, 3036,
  1972, 3287,
  1973, 1759,
  1974, 2402,
  1975, 1738,
  1976, 1010,
  1977, 2177,
  1978, 2063,
  1979, 1623,
  1980, 1730,
  1981, 1248,
  1982, 1895,
  1983, 2463,
  1984, 2276,
  1985, 3589,
  1986, 4195,
  1987, 2823,
  1988, 3450,
  1989, 4157,
  1990, 4570,
  1991, 2719,
  1992, 4083,
  1993, 6586,
  1994, 4617,
  1995, 5137,
  1996, 7796,
  1997, 6564,
  1998, 7405,
  1999, 7298,
  2000, 7867,
  2001, 7580,
  2002, 9771,
  2003, 11647,
  2004, 25827,
  2005, 25616,
  2006, 15632,
  2007, 10454,
  2008, 13278,
  2009, 16858,
  2010, 27550,
  2011, 18719,
  2012, 48277,
  2013, 28639,
  2014, 32971,
  2015, 20762,
  2016, 17972,
  2017, 18975,
  2018, 15609,
  2019, 18617,
  2020, 6124,
  2021, 2116,
  2022, 3044,
  2023, 7063
)
```



> Q1. Make a plot of `year` vs `cases` 


```{r}
library(ggplot2)
```

```{r}
library(ggplot2)

ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line()
```



> Q2. Add some major milestones including the first wP vaccine roll-out (1946), the switch  to the newer aP vaccine (1946), the switch ti the newer aP vaccine (1996), the COVID years (2020)

```{r}
ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 1946, col= "blue", lty=2) +
  geom_vline(xintercept = 1996, col="darkred", lty=2) +
  geom_vline(xintercept = 2020, col="grey", lty=2)
```

Answer: There were high case numbers in the pre 1940s; then there was a dramatic drop from of cases that time to the 1960-1970s. At this time it  it contunited to be very small ( or much smaller) amount of cases than before 1946 /1940s.

> Q3. Describe what happened after happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

In the short term, aP was able to combat Pertussis, however the vaccine was not good at doing this for a long term effect. Which is most likely what happened and is shown in the data. The immunity from aP decreases much faster (protection decreases).


**Why is this vaccine-preventable disease on the upswing?**
To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

## CMI-PB project 

[Computational Models of Immunity - Pertussis Boost](https://www.cmi-pb.org/) project aims to provide the scientific community with this very information.

They make their data available via JSON format reutrning API. We can read this in Twith the `read_json` function from the **jsonlite** package:


```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/subject", simplifyVector = TRUE)

head(subject)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```


> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```


WP is higher for on go back!!!!


> Q6. What is the breakdown of race and biological sex (e.g. number of Asian females, White males etc..)?
(In terms of race and gender is this dataset representative of the US population?)


```{r}
table(subject$race, subject$biological_sex) 
```
Answer:  There is 32 Asian females and 12 Asian males, 48 White females and 32 White males, 2 Black females and 3 Black males. Then there is 0  American Indian/Alaska Native females and 1 American Indian/Alaska Native  male. There is one female and one male Native Hawaiian or Other Pacific Islander. 15 female and 4 male individuals of More Than One Race. There are 14 females and 7 males that so not have there race reported.    

Let's read the next two: 


```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", simplifyVector = TRUE)

ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer", simplifyVector = TRUE)
```


```{r}
head(specimen)
```


```{r}
head(ab_titer)
```

To analyze this data we need to first "join" (merge/link) the different tables so we have all the data in one place not spread across different tables.


We can use the `*_join()` family of functions from **dplyr** to do this


```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
head(meta)
```


```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```
> Q. What Antibody isotypes are measured for these patients? 
Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

Answer:  IgE: 6,698 specimens
IgG: 7,265 specimens
IgG1: 11,993 specimens
IgG2: 12,000 specimens
IgG3: 12,000 specimens
IgG4: 12,000 specimens


> Q. what antigens are reported?
Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most "recent" dataset?

```{r}
table(abdata$antigen)
```
```{r}
table(abdata$dataset)
```
  
 
 
 Answer:  There is 
 2020_dataset:   31520  
2021_dataset:   8085  
2022_dataset:    7301  
2023_dataset:   15050.

In recent years there is much less rows than in earlier years ( 2020 has much more than 2023 and 2022). This may show that earlier years had much more data collected. 

 
Let's focus on the IgG antigen and make a plot of MFI_normalized for all antigens.

```{r}
igg <- abdata |>
  filter(isotype == "IgG")

head(igg)
```


```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
```

> Q 13.

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  xlim(0,75) +
  facet_wrap(vars(visit), nrow=2)
```





> Q. Is there a differenece for aP vs wP indivduals with these values?\


```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot()
facet_wrap(~infancy_vac)
```


```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot()
```

> Q. Is there a temprol response - i.e do valus increase our decrease over time?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
geom_boxplot() +
  facet_wrap(~visit)
```

## Focus on "PT" pertussis Toxin antigen

```{r}
pt.igg.21 <- igg |> filter(antigen == "PT", 
                           dataset== "2021_dataset")
```




```{r}
ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost, MFI_normalised, 
      col=infancy_vac, 
      group= subject_id) +
  geom_point() +
  geom_line()
geom_vline(xintercept = 14, col="grey", lty=2)
```


```{r, warning=FALSE}

ggplot(pt.igg.21) +
  aes(planned_day_relative_to_boost, MFI_normalised, 
      col = infancy_vac, 
      group = subject_id) +
  geom_point() +
  geom_line(alpha=0.25) +
  geom_vline(xintercept = 14, col = "grey", lty = 2) +
  geom_smooth(aes(col=infancy_vac, group=NULL),
    method = "loess",
    se = FALSE, span=0.4
  )
```







